---
title: Topics
author: Marie-Pierre Etienne
date: '2024/07/24 (updated: `r Sys.Date()`)'
institute: https://github.com/marieetienne
execute: 
  freeze: auto
---



```{r setup, include=FALSE, eval = TRUE}
with_sol <- TRUE ## in order to control the output
with_course <- TRUE
library('flipbookr')
library(RefManageR)
library(tidyverse)
library(ggplot2)
```

```{r xaringanExtra-share-again, echo=FALSE}
xaringanExtra::use_share_again()
```

```{r reference,  include=FALSE, cache=FALSE, eval = TRUE}
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "alphabetic",
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)
myBib <- ReadBib("./topics.bib", check = FALSE)
```


# Introduction 

## Linear ALgebra : What and why

::: {.callout-note icon=false}
### Definition

Linear algebra is the branch of mathematics concerning linear equations such as $a_1 x_1 + \cdots  a_n x_n = b,$ 
and linear maps such as $(x_{1},\ldots ,x_{n})\mapsto a_{1}x_{1}+\cdots +a_{n}x_{n},$
and their representations in vector spaces and through matrices.
:::

### Link between Data science and Linear Algebra
Data are organized in rectangular array, understood as matrices, and the mathematical concept associated with matrices are used to manipulate data.


### Palmer penguins example

```{r}
## install.packages('palmerpenguins")
library("palmerpenguins")
## remove_na for now
penguins_nona <- na.omit(penguins)
dim(penguins_nona)
penguins_nona[1:3,]
```

One row describes one specific penguin, while each column corresponds to one specific attribute of the penguin. Focusing on quantitative variables, we get rectangular array with `r nrow(penguins_nona)` rows, and `r sum(sapply(penguins_nona[1,], is.numeric))` columns.

```{r}
#| echo: true
penguins_nona[,sapply(penguins_nona[1,], is.numeric)]
```

Therefore, this dataset might be considered as a matrix and the linear algebra mathematical tools will be proven useful to handle dataset.


# Linear algebra - basic objects

# Vector and vector space



Based on `r Citet(myBib, "yoshida2021linear")` 

## What is Linear ALgebra



## Vectors

## Vector operations

## Some exercises

# Linear application

## Linear equation 

## System of linear equations

## Matrix operation

# Linear Algebra and Geometry

## Geometrical point of view

## Dot Product, norm and angles

## Othogonality

## Invariance : Eigen Vector, eigen values

## Change of basis 

---

```{r ggplot_back, echo = FALSE, eval = TRUE}
ggplot <- function(...) ggplot2::ggplot(...) 
```



