---
title: Linear Algebra For Data Science
author: Marie-Pierre Etienne
date: '2024/07/24 (updated: `r Sys.Date()`)'
institute: https://github.com/marieetienne
csl: ../courses_tools/resources/apa-no-doi-no-issue.csl
output:
  xaringan::moon_reader:
    css: [  'metropolis',  'mpe_pres.css']
    lib_dir: libs
    nature:
      ratio: 16:10
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: '../courses_tools/resources/collapseoutput.js'
    includes:
      after_body: '../courses_tools/resources/insert-logo.html'
fontsize: 11pt
params:
  child_path: ''
  setup_path: ../courses_tools/resources/
---



```{r setup, include=FALSE, eval = TRUE}
source(paste0(params$setup_path, "knitr_setup.R"))
with_sol <- TRUE ## in order to control the output
with_course <- TRUE
library('flipbookr')
library(RefManageR)
library(tidyverse)
library(ggplot2)
```

```{r xaringanExtra-share-again, echo=FALSE}
xaringanExtra::use_share_again()
```

```{r reference,  include=FALSE, cache=FALSE, eval = TRUE}
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "alphabetic",
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)
myBib <- ReadBib("./topics.bib", check = FALSE)
```

name: intro
# Introduction - test

## What is Linear ALgebra

## Why is Linear Algebra important in Statistics

# Vector and vector space

## Vectors

## Vector operations

## Some exercises

# Linear application

## Linear equation 

## System of linear equations

## Matrix operation

# Linear Algebra and Geometry

## Geometrical point of view

## Dot Product, norm and angles

## Othogonality

## Invariance : Eigen Vector, eigen values

## Change of basis 

---

```{r ggplot_back, echo = FALSE, eval = TRUE}
ggplot <- function(...) ggplot2::ggplot(...) 
```



